---
title: "Introduction to Open Data Science"
author: "Gyan Dookie"
email: "gyan.dookie@gmail.com"
date: "8 March 2017"
output: 
  html_document:
   theme: cosmo
   toc: true
   toc_depth: 2
   fig_caption: true
   fig_width: 6
   fig_height: 4
   code_folding: hide
---

## **Final assignment - Gyan Dookie**

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# knitr::opts_chunk$set(echo = TRUE)


#install.packages("dplyr",  repos = "https://ftp.acc.umu.se/mirror/CRAN/")
#install.packages("corrplot")

#install.packages("tidyr")
library(tidyr)
library(dplyr)
library(corrplot)
library(tidyr)

library(ggplot2)
library(GGally)


setwd("/Users/gyandookie/IODS-final")


refBomb<- read.csv ("data/refBomb.csv", header=TRUE, sep=",", stringsAsFactors = FALSE)


```

#Final Assignment
### Overview
##### We live in a turbulent world today. One current socially destabilizing and publically debated phenomena has been the flow of refugees and asylum seekers from Middle East to Europe. In this IODS final assignment my aim is to explore the possible relations among variables that are in my view connected to the ongoing so called “refugee crisis”. I’ll be focusing on countries that are bombing the countries where the major flow of refugees originate from. Here are some questions that interest me. How many refugees do the “bombing” countries have / accept inside their borders? Does taking part in foreign wars increase the risk of terror attacks? How are these warring countries connected to the arms trade? How succesful economically and socially are these countries? My hypothesis is at this point still vague. It could be put bluntly though: “Rich countries that are main players in the arms business are bombing poor countries and by doing so are increasing refugee flows across their borders and the risk of terrorist attacks.” Nevertheless, this would be likely oversimplifying the reality.


 In the following data analysis I’m going to  perform Principal Component Analysis (PCA) on the ref dataset. Before the actual analysis I'll go through some preliminary explorations of the data.
 
 #### Data wrangling
 
 ##### Variables

It's a good idea to introduce the variables of the ref dataset before moving further.

* Country
* FSI - Fragile States Index, a variable combined of other variables estimating e.g. stability, human rights, freedom of press in a country
    * This data is from Fund for Peace's yearly published  Fragile States -dataset
* TerAtt = How many islamist terror attacks has there been in the country between 2015-2017
* CtryBomb = Hown many countries the country has been bombing between 2015-2016
* ArmsSale = The amount of a country's arms exports (2015)
    * This data is from SIPRI's MILEX dataset
* RefOrig = The amount of refugees originating from a country
* GNI = Gross National Income per Capita
     * Gross National Income data was provided by World Bank
* refPer - The amount of refugees in the asylum country
    * This variable was produced/mutated by me (by  multiplying the amount of refugees / asylum seekers in a country of asylum by hundred and  then dividing it with the population of that country)
 
##### Methods
The data was selected from multiple sources as can seen above. My aim was to bring together data related to the broader so called refugee crisis -phenoma, but that hasn't necessary been so often brought together. I ventured to also add some variables that in my view are actants in this larger phenomena. I had to perform many different data wrangling operations to combine all the data into one dataset in a refined form. Here are some of the methods I used: importing, selecting, mutating, deleting or completing data rows with blank (e.g. NA) values, changing the datatype of variables, combining data.

    

Here is the link to my data wrangling file.

* [The datawrangling file](https://github.com/gdookie/IODS-final/blob/master/data/create_ref.R)

### Brief description of the used methods

My method is to bring together different variables that seem to revolve around a phenomena and see what kind of correlations this operation reveals. After some basic explorations (i.e. the structure and summaries of the data) I’ll produce the correlations of the data and plot these visually. In the next phase I’ll use the Principal component analysis (PCA) and interpret what this reveals. Finally I’ll add some more visualizations of the data.


 

### 1. Preliminary explorations of the data
#### **1.1 The structure and the dimensions of the data**
Below is the structure of the human dataset. The following characteristics of the dataframe can be discerned.

* 155 rows
* 8 variables



```{r, echo=TRUE}
refBomb<- refBomb[-1]
dim(refBomb)
str(refBomb)

```
<!--Showing a graphical overview of the data and showing summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. (0-2 points)-->

#### **1.2 The summary of the data**
Here we'll print out the summary of the data with the *summary()* function to get a grasp of the min, max, median, mean and quantiles of the data. 

```{r, echo=TRUE}
summary(refBomb)

```

Here are the standard deviations of the variables.


```{r, echo=TRUE}
# Lasketaan muuttujien keskihajonnat (tehdään tämä uudestaan, kun muuttujat on skaalattu)
summarise (refBomb, sd(FSI),sd(TerAtt),sd(CtryBomb),sd(ArmSale),sd(RefOrig),sd(GNI),sd(refPer))



```

#### **1.3 The graphical overview of the summarized data**
Let's visualize our data to get a better overall picture of it. First we'll produce a matrix plot with the basic package's pairs() and then with GGally package's ggpairs().

```{r, echo=TRUE}
# Then we'll take a look at the graphical overview
refBomb_1 <- select(refBomb, -Country)
pairs(refBomb_1)
ggpairs(refBomb_1)



```

#### **1.4 The correlations of the data with corrplot()**

Now it's time to produce a table of the correlations with the  *cor()* function. Here the correlations were rounded to two desimals to save space.

```{r, echo=TRUE}

cor_matrix<-cor(refBomb_1)%>% round(2)
cor_matrix


```

#### **1.5 The graphical overview of correlations with the advanced corrplot() function**

Here's the visualization of the correlation matrix with the advanced *corrplot()* function. To reduce repetition, we'll visualize only the upper part of the plot (as is well known, the top part of the correlation matrix contains the same correlations as the bottom part)

```{r, echo=TRUE}
# Now we'll visualize the correlation matrix. To reduce repetition, we'll visualize only the upper part of the plot (as is well known, the top part of the correlation matrix contains the same correlations as the bottom part)
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos = "d", tl.cex=0.6)


```

#### **1.6 The above summaries and visualizations showed the following**

**Here are some of the exposed correlations**

* Strong positive correlations between variable pairs
    * There were relatively few strong correlations between variables
    * A strong correlation showed between Military arms exporting and bombing other countries. From this we can deduce that countries selling arms are more prone to bomb other countries. One intepretation of this would be that it is in the interest of arms selling countries to test their weapons in the battle field. The battle filed can also be seen as a place to advertise the goods to possible onlooking customers.
    * A strong correlation between the percentage of refugees (refPer) in an asylum country and the amount of refugees originating in a country was not surprising. The heaviest refugee load is usually beared by the countries in conflict.
    * GNI and the countries Fragility Index had a negative connection. The richer the country and it's citizens, the more stabile that country is likely to be.
* Moderate positive correlation between the following variable pairs
    * A noticeable, but not so strong correlation showed between how much the country bombed other countries and Islamist terrorist attacks taking place in that country. My hypothesis was originally that this "revenge" connection would be stronger. One aspect that could weaken this connection is that islamist terrorist attacks happened in the conflict hotspot countries (i.e. Afganistan, Pakistan, Iraq). Some of these could be categorized as resistance in a civil war.
* Another correlation of similar strength was found between GNI and bombing other countries. This finding supports my hypothesis of rich countries bombing poor countries.

**The distributions of the data**

* Allmost all the variables were (left) skewed
* The FSI variables distribution was closest to a normal distribution and had a slight bimodal tendency

<!--  More plots here (boxplots, density plots, bar plots) -->




```{r, echo=TRUE}
# 
# boxplots

```



```{r, echo=TRUE}

```



```{r, echo=TRUE}

```

    


```

Then we'll draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. (0-2 points)
```{r, echo=TRUE, message= FALSE, warning=FALSE}

# bibplot of first two principal components
# draw a biplot of the principal component representation and the original variables
# biplot(pca_human, choices = 1:2, cex=c(1.0 ,1.0), col= c("grey40", "blue"), xlab = pc_lab[1] , ylab = pc_lab[2])

```



#### **2.2 PCA on standardized/scaled data**
Here are the summaries of the scaled variables. See how the variables changed ( e.g. the means are now all at zero). As we can see below, the distribution of explainability is now more spread among the PC's. The PCA plot also makes now a lot more sense.


```{r, echo=TRUE, warning=FALSE, message=FALSE}
# standardize the variables
refBomb_std <- scale(refBomb_1)
summary(refBomb_std)
#str(human_std)

```
```{r, echo=FALSE}

# Lasketaan standardoitujen muuttujien keskihajonnat
#summarise(human_std_mat, sd(Edu2.FM),sd(Labo.FM),sd(Edu.Exp),sd(Life.Exp),sd(GNI),sd(Mat.Mor),sd(Ado.Birth),sd(Parli.F)

```
```{r, echo=TRUE}

# perform principal component analysis (with the SVD method)
pca_refBomb <- prcomp(refBomb_std)
s <- summary(pca_refBomb)
pca_pr <- round(100*s$importance[2, ], digits = 1)
pca_pr
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot of the principal component representation and the original variables
## remember to add caption names (in place of variable names) to the plots describing the phenomenons


#biplot(pca_human, choices = 1:3, cex=c(1 ,1.2), col= c("grey40", "deeppink2"))

biplot(pca_refBomb, choices = 1:2, cex = c(0.8, 0.8), col = c("black", "red"), xlab = pc_lab[1] , ylab = pc_lab[2], main="\n Arms sales, bombing foreign countries, refugees and \n the stability/fragility of a country. \n\n\n")
```

**Let's take a closer look at the countries on the "west side" of the biplot and close to PC2.**

```{r,echo=TRUE}

biplot(pca_refBomb, expand=4, xlim=c(0.5, -0.2), ylim=c(-0.1, 0.1))
```



#### **2.3 Intepretation and analysis of PCA and the corresponding biplots**
The PCA biplot based on the standardized variables offers visual information that supports our earlier findings.






```{r}

```

**Intepretation of PC1**

Generally speaking, the 1st principal component captures the maximum amount of variance from the features in the original data. Here the amount of variance of the data captured by PC1 is 32.4 %.
The variables/features connected to the PC1 dimension are CtryBomb, ArmSale and TerAtt which point their arrows in a close cluster and in a slight angle horizontally to the right. So from this we can deduce that these variables have some correlation among each other.
 FSI and GNI are pointing to oppisite directions and thus are negatively correlated. The FSI-GNI -arrowline is in a slight angle. It is more aligned with the PC1 thatn PC2.


**Intepretation of PC2**

The 2nd principal component PC2 is orthogonal to the first and it captures the maximum amount of variability/variance left. Here that amount is 27.2 % and variables in the closest relationship to this principal component are RefOrig and perP. So the stability of a country is connected to the amount of refugees in that country.





```{r, echo=FALSE, message=FALSE}

```



```{r, echo=FALSE, warning=FALSE}


```

<!-- #### **3.2 The structure and the summary of the subsetted data** -->

```{r, echo=FALSE}



```

<!-- #### **3.3 The visual overview of the data** -->

```{r, echo=FALSE, warning=FALSE}

# # visualize the dataset
# gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))


```



```{r, echo=FALSE}
#summary(tea)

```




````{r, echo=FALSE}




```



```{r, echo=FALSE, warning=FALSE}


```


```{r, echo=FALSE}
# #names(mca)
# # visualize MCA
# plot(mca, invisible=c(habillage="quali"), title= "MCA factor map - tea drinking vs. \n sport, social and spiritual orientations")




```



```{r, echo=FALSE}

# ggplot(tea_time, aes(x = sex, fill = friends)) + 
#   geom_bar(position = "fill") + ylab("Proportion")

```




```{r,echo=FALSE}

# ggplot(tea_time, aes(x = sex, fill = sophisticated)) + 
#   geom_bar(position = "fill") + ylab("Proportion")

```



```{r, echo=FALSE}

```

```{r,echo=FALSE}

```



```{r, echo=FALSE}

```



```{r,echo=FALSE}


```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

