---
title: "Introduction to Open Data Science"
author: "Gyan Dookie"
email: "gyan.dookie@gmail.com"
date: "8 March 2017"
output: html_document
theme: cosmo
toc: true
toc_depth: 2
fig_caption: true
fig_width: 6
fig_height: 4
code_folding: hide
---

## **Final assignment - Gyan Dookie**

```{r, echo=FALSE, message=FALSE, warning=FALSE}


#install.packages("dplyr",  repos = "https://ftp.acc.umu.se/mirror/CRAN/")
#install.packages("corrplot")

#install.packages("tidyr")
library(tidyr)
library(dplyr)
library(corrplot)
library(tidyr)

library(ggplot2)
library(GGally)

# Load the ‘human’ data into R. Explore the structure and the dimensions of the data and describe the dataset briefly, assuming the reader has no previous knowledge of it (this is now close to the reality, since you have named the variables yourself). (0-1 point)

#human <- read.table("data/human.csv", header= TRUE, sep=",")
setwd("/Users/gyandookie/IODS-final")

#refBomb<- read.csv ("data/refBomb.csv", header=TRUE, sep=",", quote="", stringsAsFactors = FALSE)
refBomb<- read.csv ("data/refBomb.csv", header=TRUE, sep=",", stringsAsFactors = FALSE)


```

#Final Assignment
### Overview
##### We live in a turbulent world today. One current socially destabilizing and publically debated phenomena has been the flow of refugees and asylum seekers from Middle East to Europe. In this IODS final assignment my aim is to explore the possible relations among variables that are in my view connected to the ongoing so called “refugee crisis”. I’ll be focusing on countries that are bombing the countries where the major flow of refugees originate from. Here are some questions that interest me. How many refugees do the “bombing” countries have / accept inside their borders? Does taking part in foreign wars increase the risk of terror attacks? How are these warring countries connected to the arms trade? How succesful economically and socially are these countries? My hypothesis is at this point still vague. It could be put bluntly though: “Rich countries that are main players in the arms business are bombing poor countries and by doing so are increasing refugee flows across their borders and the risk of terrorist attacks.” Nevertheless, this would be likely oversimplifying the reality.


 In the following data analysis I’m going to  perform Principal Component Analysis (PCA) on the ref dataset. Before the actual analysis I'll go through some preliminary explorations of the data.
 
 #### Data wrangling
 
 ##### Variables

It's a good idea to introduce the variables of the ref dataset before moving further.

* Country
* FSI - Fragile States Index, a variable combined of other variables estimating e.g. stability, human rights, freedom of press in a country
    * This data is from Fund for Peace's yearly published  Fragile States -dataset
* TerAtt = How many islamist terror attacks has there been in the country between 2015-2017
* CtryBomb = Hown many countries the country has been bombing between 2015-2016
* ArmsSale = The amount of a country's arms exports (2015)
    * This data is from SIPRI's MILEX dataset
* RefOrig = The amount of refugees originating from a country
* GNI = Gross National Income per Capita
     * Gross National Income data was provided by World Bank
* refPer - The amount of refugees in the asylum country
    * This variable was produced/mutated by me (by  multiplying the amount of refugees / asylum seekers in a country of asylum by hundred and  then dividing it with the population of that country)
 
##### Methods
The data was selected from multiple sources as can seen above. My aim was to bring together data related to the broader so called refugee crisis -phenoma, but that hasn't necessary been so often brought together. I ventured to also add some variables that in my view are actants in this larger phenomena. I had to perform many different data wrangling operations to combine all the data into one dataset in a refined form. Here are some of the methods I used: importing, selecting, mutating, deleting or completing data rows with blank (e.g. NA) values, changing the datatype of variables, combining data.

    

Here is the link to my data wrangling file.

* [The datawrangling file](https://github.com/gdookie/IODS-final/blob/master/data/create_ref.R)

### Brief description of the used methods

My method is to bring together different variables that seem to revolve around a phenomena and see what kind of correlations this operation reveals. After some basic explorations (i.e. the structure and summaries of the data) I’ll produce the correlations of the data and plot these visually. In the next phase I’ll use the Principal component analysis (PCA) and interpret what this reveals. Finally I’ll add some more visualizations of the data.


 

### 1. Preliminary explorations of the data
#### **1.1 The structure and the dimensions of the data**
Below is the structure of the human dataset. The following characteristics of the dataframe can be discerned.

* 155 rows
* 8 variables



```{r, echo=FALSE}
refBomb<- refBomb[-1]
dim(refBomb)
str(refBomb)

```
<!--Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. (0-2 points)-->

#### **1.2 The summary of the data**
Here we'll print out the summary of the data with the *summary()* function to get a grasp of the min, max, median, mean and quantiles of the data. 

```{r, echo=FALSE}
summary(refBomb)

```

Here are the standard deviations of the variables.


```{r, echo=FALSE}
# Lasketaan muuttujien keskihajonnat (tehdään tämä uudestaan, kun muuttujat on skaalattu)
summarise (refBomb, sd(FSI),sd(TerAtt),sd(CtryBomb),sd(ArmSale),sd(RefOrig),sd(GNI),sd(refPer))



```

#### **1.3 The graphical overview of the summarized data**
Let's visualize our data to get a better overall picture of it. First we'll produce a matrix plot with the basic package's pairs() and then with GGally package's ggpairs().

```{r, echo=FALSE}
# Then we'll take a look at the graphical overview
refBomb_1 <- select(refBomb, -Country)
pairs(refBomb_1)
ggpairs(refBomb_1)



```

#### **1.4 The correlations of the data with corrplot()**

Now it's time to produce a table of the correlations with the  *cor()* function. Here the correlations were rounded to two desimals to save space.

```{r, echo=FALSE}

cor_matrix<-cor(refBomb_1)%>% round(2)
cor_matrix


```

#### **1.5 The graphical overview of correlations with the advanced corrplot() function**

Here's the visualization of the correlation matrix with the advanced *corrplot()* function. To reduce repetition, we'll visualize only the upper part of the plot (as is well known, the top part of the correlation matrix contains the same correlations as the bottom part)

```{r, echo=FALSE}
# Now we'll visualize the correlation matrix. To reduce repetition, we'll visualize only the upper part of the plot (as is well known, the top part of the correlation matrix contains the same correlations as the bottom part)
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos = "d", tl.cex=0.6)


```

#### **1.6 The above summaries and visualizations showed the following**

**Here are some of the exposed correlations**

* Strong positive correlations between variable pairs
    * There were relatively few strong correlations between variables
    * A strong correlation showed between Military arms exporting and bombing other countries. From this we can deduce that countries selling arms are more prone to bomb other countries. One intepretation of this would be that it is in the interest of arms selling countries to test their weapons in the battle field. The battle filed can also be seen as a place to advertise the goods to possible onlooking customers.
    * A strong correlation between the percentage of refugees (refPer) in an asylum country and the amount of refugees originating in a country was not surprising. The heaviest refugee load is usually beared by the countries in conflict.
    * GNI and the countries Fragility Index had a negative connection. The richer the country and it's citizens, the more stabile that country is likely to be.
* Moderate positive correlation between the following variable pairs
    * A noticeable, but not so strong correlation showed between how much the country bombed other countries and Islamist terrorist attacks taking place in that country. My hypothesis was originally that this "revenge" connection would be stronger. One aspect that could weaken this connection is that islamist terrorist attacks happened in the conflict hotspot countries (i.e. Afganistan, Pakistan, Iraq). Some of these could be categorized as resistance in a civil war.
* Another correlation of similar strength was found between GNI and bombing other countries. This finding supports my hypothesis of rich countries bombing poor countries.



<!--  More plots here (boxplots, density plots, bar plots) -->




```{r, echo=FALSE}
# 
# boxplots

```



```{r, echo=FALSE}

```



```{r, echo=FALSE}
# library(dplyr)
# 
#  boston_scaled %>%
#   summarise(sd(crim),
#             sd(zn), sd(indus), sd(chas), sd(nox), sd(rm), sd(age), sd(dis), sd(rad), sd(tax),sd(ptratio), sd(black), sd(lstat), sd(medv))
```

    
<!-- ### **2. Performing the Principal component analysis (PCA) on the human data** -->
<!-- #### **2.1 PCA on non-standardized data** -->
<!-- Next we'll perform principal component analysis (PCA) on the not standardized human data and show variability captured by the principal components. -->
```{r, echo=FALSE, message= FALSE, warning=FALSE}
### PCA on the non standardized human data
# pca_human <- prcomp(human)
# s <- summary(pca_human)
# 
# pca_pr <- round(100*s$importance[2, ], digits = 1)
# pca_pr
# pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

## puuttuuko jotain dataa, kun ilmoittaa zero-lenght

```

Then we'll draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. (0-2 points)
```{r, echo=FALSE, message= FALSE, warning=FALSE}

# bibplot of first two principal components
# draw a biplot of the principal component representation and the original variables
# biplot(pca_human, choices = 1:2, cex=c(1.0 ,1.0), col= c("grey40", "blue"), xlab = pc_lab[1] , ylab = pc_lab[2])

```



#### **2.2 PCA on standardized/scaled data**
Here are the summaries of the scaled variables. See how the variables changed ( e.g. the means are now all at zero). As we can see below, the distribution of explainability is now more spread among the PC's. The PCA plot also makes now a lot more sense.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
# standardize the variables
refBomb_std <- scale(refBomb_1)
summary(refBomb_std)
#str(human_std)

```
```{r, echo=FALSE}

# Lasketaan standardoitujen muuttujien keskihajonnat
#summarise(human_std_mat, sd(Edu2.FM),sd(Labo.FM),sd(Edu.Exp),sd(Life.Exp),sd(GNI),sd(Mat.Mor),sd(Ado.Birth),sd(Parli.F)

```
```{r, echo=FALSE}

# perform principal component analysis (with the SVD method)
pca_refBomb <- prcomp(refBomb_std)
s <- summary(pca_refBomb)
pca_pr <- round(100*s$importance[2, ], digits = 1)
pca_pr
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot of the principal component representation and the original variables
## remember to add caption names (in place of variable names) to the plots describing the phenomenons


#biplot(pca_human, choices = 1:3, cex=c(1 ,1.2), col= c("grey40", "deeppink2"))

biplot(pca_refBomb, choices = 1:2, cex = c(0.8, 0.8), col = c("black", "red"), xlab = pc_lab[1] , ylab = pc_lab[2], main="\n Arms sales, bombing, refugees and \n the stability/fragility of a country. \n\n\n")
```

**Let's take a closer look at the countries on the "west side" of the biplot and close to PC2.**

```{r,echo=FALSE}

biplot(pca_refBomb, expand=4, xlim=c(0.5, -0.2), ylim=c(-0.1, 0.1))
```

<!-- Let's interpret the results of both analysis (with and without standardizing). Are the results different? Why or why not? Including captions in the plots where I describe the results by using not just your variable names, but the actual phenomenons they relate to. (0-4 points) -->

<!-- Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data. (0-2 points) -->

#### **2.3 Intepretation and analysis of PCA and the corresponding biplots**
The PCA biplot based on the standardized variables offers visual information that supports our earlier findings.

*




```{r}

```

**Intepretation of PC1**

Generally speaking, the 1st principal component captures the maximum amount of variance from the features in the original data. Here the amount of variance of the data captured by PC1 is 32.4 %.
The variables/features connected to the PC1 dimension are CtryBomb, ArmSale and TerAtt which point their arrows in a close cluster and in a slight angle horizontally to the right. So from this we can deduce that these variables have some correlation among each other.
 FSI and GNI are pointing to oppisite directions and thus are negatively correlated. The FSI-GNI -arrowline is in a slight angle. It is more aligned with the PC1 thatn PC2.


**Intepretation of PC2**

The 2nd principal component PC2 is orthogonal to the first and it captures the maximum amount of variability/variance left. Here that amount is 27.2 % and variables in the closest relationship to this principal component are RefOrig and perP. So the stability of a country is connected to the amount of refugees in that country.


<!-- ### **3. Performing the Multiple Correspondence Analysis (MCA) on the tea data** -->

<!-- Next we'll load the tea dataset from the package Factominer and explore the data briefly. -->


```{r, echo=FALSE, message=FALSE}
# # Install Factominer and load in the tea dataset
# #install.packages("FactoMineR")
# library(FactoMineR)
# data(tea)

```

<!-- #### **3.1 The structure and dimensions of the data** -->

<!-- Let's look at the structure and the dimensions of the data first. Then we'll create a subset of it by selecting the following variables. -->

<!-- * Sport -->
<!-- * effect.on.health -->
<!-- * sophisticated -->
<!-- * spirituality -->
<!-- * friends -->
<!-- * sex -->

```{r, echo=FALSE, warning=FALSE}
# # column names to keep in the dataset
# keep_columns <- c("Sport", "effect.on.health", "sophisticated", "spirituality", "friends", "sex")
# #keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
# keep_columns
# #head(tea)
# dim(tea)
# str(tea)
# names(tea)
# # select the 'keep_columns' to create a new dataset
# #tea <- data.frame(tea)
# tea_time <- dplyr::select(tea, one_of(keep_columns))

```

<!-- #### **3.2 The structure and the summary of the subsetted data** -->

```{r, echo=FALSE}

# # look at the summaries and structure of the data
# str(tea_time)
# summary(tea_time)


```

<!-- #### **3.3 The visual overview of the data** -->

```{r, echo=FALSE, warning=FALSE}

# # visualize the dataset
# gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))


```

<!-- Here's a summary of the tea data's variables and the distributions within their levels. -->

```{r, echo=FALSE}
#summary(tea)

```

<!-- Let's take a look at the correlations of the tea dataset with cor() (produces a correlation matrix) and corrplot() (visualizes the correlation matrix as pairs)-->


````{r, echo=FALSE}



# Now we'll visualize the correlation matrix. To reduce repetition, we'll visualize only the upper part of the plot (as is well known, the top part of the correlation matrix contains the same correlations as the bottom part)
#corrplot(cor_matrixT, method="circle", type="upper", cl.pos="b", tl.pos = "d", tl.cex=0.6)
```

<!-- #### **3.4 Performing the Multiple Correspondence analysis on the data** -->
<!-- Let's do the multiple correspondence analysis of selected tea variables. -->

```{r, echo=FALSE, warning=FALSE}

# # multiple correspondence analysis
# mca <- MCA(tea_time, graph = FALSE)
# 
# # summary of the model
# summary(mca)
```

<!-- #### **3.5 Analyzing the MCA's summary table** -->

<!-- * The eigen values show the amount of variance captured by the different dimensions. Here we can see the following. -->
<!--     * Dimension 1 captures 22.5 % of the variance within the data  -->
<!--     * Dimension 2 captures 17.5 % of the variance within the data -->
<!--     *The first two dimensions capture approximately 40 % of the variance within the data. This amount is considerably lower than the amount captured by the first two principal components in the previous analysis of the human dataset. -->
<!-- * The individuals -->
<!--     * the individuals coordinates, the individuals contribution (%) on the dimension and the cos2 (the squared correlations) on the dimensions. -->
<!-- * The categories -->
<!--     * The Categories part shows coordinates of the variable categories, the contribution (%), the cos2 (the squared correlations) and v.test value. -->

<!-- * The Categorical value -->
<!--     * Shows the squared correlation between each variable and the dimensions -->
<!--     * If the value is close to one it indicates a strong link with the variable and dimension. Here (among the three dimensions shown), only Dimension 3 and and the effect.on.health have a categorical variable that is close to one (0.876) -->


<!-- #### **3.6 Visualizing the Multiple Correspondence Analysis with the plot() function** -->
```{r, echo=FALSE}
# #names(mca)
# # visualize MCA
# plot(mca, invisible=c(habillage="quali"), title= "MCA factor map - tea drinking vs. \n sport, social and spiritual orientations")




```

<!-- #### **3.7 Some conclusions of the MCA and it's plot** -->
<!-- * There’s a clear difference in the social role of tea between females and males. Men prefer drinking tea alone more often than women and women tend to drink tea with friends more often than men. -->

<!-- This proportional barplot confirms that women drink tea more with friends than men do (which was also suggested by the MCA-plot above). -->

```{r, echo=FALSE}

# ggplot(tea_time, aes(x = sex, fill = friends)) + 
#   geom_bar(position = "fill") + ylab("Proportion")

```


<!-- The above MCA-biplot showed that women regard tea drinking more than men as sophisticated. This finding is confirmed in the barplot below. -->

```{r,echo=FALSE}

# ggplot(tea_time, aes(x = sex, fill = sophisticated)) + 
#   geom_bar(position = "fill") + ylab("Proportion")

```



```{r, echo=FALSE}

```

```{r,echo=FALSE}

```



```{r, echo=FALSE}

```



```{r,echo=FALSE}


```

***

<!-- ```{r child = "final.Rmd"} -->
<!-- ``` -->
  
***

***
  
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

